---
title: "Harmonize sex information in cbio data"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```


# Overview

This .qmd file demonstrates a workflow for curating and harmonizing 
sex data with the cbioportal dataset.
This file curates information from 6 original columns to produce a curated 
column relating to the sex of the patient.


# Setup

## Load Packages

```{r load}
suppressPackageStartupMessages({
 library(dplyr)
library(tidyverse)
library(stringr)
library(rvest)
library(rols)
library(vctrs)
library(hash)
})
```


## Setup for Curation

We will first add a `curation_id` consisting of study ID, patient ID, and sample ID (`studyId:patientId:sampleId`) to avoid confusion due to duplicated samples.
```{r curation_setup}
# Load cbio data
cbio <- readRDS("D:\\CUNY Internship\\cBioPortal_all_clinicalData_combined_2023-05-18.rds")

# Add curation ID
cbio$curation_id <- paste(cbio$studyId, cbio$patientId, cbio$sampleId, sep = ":")

# Setup google sheets link to mapping table
url <- "https://docs.google.com/spreadsheets/d/1O_W_QDUZKWRNy4GbDQp3RCObb0FCBSVZPrkQA-0zdPk/edit#gid=0"

ss <- googledrive::as_id(url)

# Define a project directory
proj_dir <- "D:\\CUNY Internship\\cbio\\Sex\\"
```


# Exploratory Data Analysis & Data Curation

## Identifying Relevant Columns & Completeness

Next, we will identify the columns containing values relating to the sex of the 
patient.
Due to the large number of feature columns which contain a number of both broad
and very specific column types, we will first query column names for relevant 
terms and compile these results into a map of original columns and their
potential curated columns.
```{r}
# Create a list of potentially relevant columns based on search terms
sex_columns <- select(cbio, contains(c("sex", "gender", "male", "genital", "cervi", "testi", "teste", "peni", "vagin", "uter", "prostat", "endome", "vulv", "fallo", "ovar")))

potential_columns <- data.frame(colnames(sex_columns))

# Iterate through these columns to create a dataframe of each column's unique values and completeness
for (col in 1:length(sex_columns)){
  potential_columns$unique_vals[col] <- paste(unlist(unique(sex_columns[,col]), use.names = F), collapse=" ")
  potential_columns$completeness[col] <- (nrow(sex_columns) - length(which(is.na(sex_columns[,col]))))/nrow(sex_columns)
}

# Remove rows where unique_vals is NA
potential_columns <- filter(potential_columns, unique_vals!="NA")

# Sort by completeness
potential_columns <- arrange(potential_columns, desc(completeness))

# Export to csv file to manually remove irrelevant columns
write.csv(potential_columns, 
          file = file.path(proj_dir, "potential_sex_columns_map.csv"),
          row.names = FALSE)
```

After manually searching for character and boolean columns with titles pertaining 
to patient sex and adding them to the map, as well as manually removing and 
rearranging some columns which were irrelevant to this particular curation or 
redundant relative to the pre-existing "sex" column, we can then create a tibble 
dataframe and import the map of columns to include in this curation:
```{r}
# Create tibble
cbio <- tibble(cbio)

# Load Potential columns map csv file
pot_col_map <- read.csv(file=file.path(proj_dir, "cBioPortal_sex_columns_map.csv"))

# Create a subset of cbio containing only relevant columns
sex_cols_cbio <- cbio[,c("curation_id", pot_col_map$colnames.sex_columns.)]
```

Before we condense these values into a single merged column, we will need to 
convert some columns from TRUE/FALSE and Yes/No values into Female/Male values:
```{r}
# Convert Yes/No and T/F columns to descriptive values
# Create a subset of female anatomy indicator columns to process together
F_indicators <- c("DMETS_DX_OVARY", "DMETS_DX_FEMALE_GENITAL", "METS_SITES_AT_METS_ABST_OVARY", "METS_SITES_AT_METS_DX_OVARY", "DIAGNOSIS_CANCER_LOC_OVARY", "ANYTIME_CANCER_LOC_OVARY")
M_indicators <- c("TESTICULAR_INVOLVEMENT", "DMETS_DX_MALE_GENITAL", "FAMILY_HISTORY_TESTICULAR_CANCER", "PRD_PROSTATECTOMY", "TESTIS_TUMOR_MACROEXTENT", "HISTORY_OF_UNDESCENDED_TESTIS", "INCIDENTAL_PROSTATE_CANCER", "TESTES_SITE_OF_RELAPSE")

# Replace positive indicator values with appropriate Female/Male value
sex_cols_cbio[,F_indicators] <- data.frame(lapply(sex_cols_cbio[,F_indicators], function(x) gsub("^Yes$", "Female", x)))
sex_cols_cbio[,F_indicators] <- data.frame(lapply(sex_cols_cbio[,F_indicators], function(x) gsub("^TRUE$", "Female", x)))
sex_cols_cbio[,F_indicators] <- data.frame(lapply(sex_cols_cbio[,F_indicators], function(x) gsub("^YES$", "Female", x)))
sex_cols_cbio[,M_indicators] <- data.frame(lapply(sex_cols_cbio[,M_indicators], function(x) gsub("^Yes$", "Male", x)))
sex_cols_cbio[,M_indicators] <- data.frame(lapply(sex_cols_cbio[,M_indicators], function(x) gsub("^YES$", "Male", x)))
sex_cols_cbio[,M_indicators] <- data.frame(lapply(sex_cols_cbio[,M_indicators], function(x) gsub("^TRUE$", "Male", x)))
sex_cols_cbio[,M_indicators] <- data.frame(lapply(sex_cols_cbio[,M_indicators], function(x) ifelse(grepl("^Yes;", x), "Male", x)))
sex_cols_cbio[,M_indicators] <- data.frame(lapply(sex_cols_cbio[,M_indicators], function(x) ifelse(grepl("^Involves", x), "Male", x)))

# Replace negative indicator values with appropriate Female/Male value
sex_cols_cbio[,c(F_indicators, M_indicators)] <- data.frame(lapply(sex_cols_cbio[,c(F_indicators, M_indicators)], function(x) gsub("^No$", NA, x)))
sex_cols_cbio[,c(F_indicators, M_indicators)] <- data.frame(lapply(sex_cols_cbio[,c(F_indicators, M_indicators)], function(x) gsub("^NO$", NA, x)))
sex_cols_cbio[,c(F_indicators, M_indicators)] <- data.frame(lapply(sex_cols_cbio[,c(F_indicators, M_indicators)], function(x) gsub("^FALSE$", NA, x)))

# Standardize capitalization
sex_cols_cbio <- data.frame(lapply(sex_cols_cbio, function(x) gsub("FEMALE", "Female", x)))
sex_cols_cbio <- data.frame(lapply(sex_cols_cbio, function(x) gsub("^MALE$", "Male", x)))

# Convert numeric values to "Female" for "CERVIX_SUV_RESULTS" column
sex_cols_cbio$CERVIX_SUV_RESULTS <- as.numeric(sex_cols_cbio$CERVIX_SUV_RESULTS)
sex_cols_cbio$CERVIX_SUV_RESULTS[sex_cols_cbio$CERVIX_SUV_RESULTS >= 0] <- "Female"

# Remove all unclassified values
approved_vals <- c("Male", "Female")
sex_cols_cbio[,c(F_indicators, M_indicators)] <- data.frame(lapply(sex_cols_cbio[,c(F_indicators, M_indicators)], function(x) ifelse(x %in% approved_vals, x, NA)))

```

Now that all columns include descriptive values, we can merge these values into 
a single column with each distinct value separated by a semicolon.
```{r}
# Merge all relevant rows
for (i in 1:nrow(sex_cols_cbio)){
  sex_cols_cbio$merged[i] <- toupper(paste(na.omit(as.vector(unlist(sex_cols_cbio[i,2:18]))), collapse="<;>"))
}
```

Then we can create a list of all unique values included in the merged column,
and export these values to a csv file to manually develop an ontology map:
```{r}
# Count the number of unique values in the merged column
merged_vals <- list()
for(r in 1:nrow(sex_cols_cbio)){
  merged_vals <- append(merged_vals, str_split(sex_cols_cbio$merged[r], ";"))
}
unique_merged_vals <- unique(unlist(merged_vals))

# Make a csv file of the unique values in the merged column
write.csv(unique_merged_vals, 
          file= file.path(proj_dir, "sex_cols_cbio_unique_vals.csv"), 
          row.names=F, col.names=F)
```


# Mapping

## Import Maps

To access the necessary sex ontology terms we will load the `sex_ontology_map`.
```{r import_maps, eval=FALSE}
# import bodysite ontology map
mapped_terms <- read_sheet(ss, sheet = "cBioPortal_sex_map")

# import bodysite ontology map
mapped_terms <-read.csv(file = file.path(proj_dir, "cBioPortal_sex_map.csv"), header=T)
```


# Curating Sex Columns

Now we can map the values in the merged columns to their corresponding ontology
terms and propagate these values into the curated columns: 
```{r}
# Make a function to remove excess semicolons
rmv_xtra_semis <- function(x){
  x <- ifelse(startsWith(x, ";"), sub("^;", "", x), x)
  return(ifelse(endsWith(x, ";"), sub(";$", "", x), x))
}

# Iterate through merged column values
for (x in 1:length(sex_cols_cbio$merged)){
  # Create a list of terms in the value
  original_terms <- as.list(unlist(strsplit(sex_cols_cbio$merged[x], ";")))
  new_terms <- list()
  new_term_ids <- list()
  # Search for replacement terms in the ontology map
  new_terms <- lapply(original_terms, function(y) mapped_terms$curated_ontology[grep(paste("^",y,"$",sep=""), mapped_terms$original_value, fixed=F)])
  new_term_ids <- lapply(original_terms, function(y) mapped_terms$curated_ontology_term_id[grep(paste("^",y,"$",sep=""), mapped_terms$original_value, fixed=F)])
  new_terms <- list_drop_empty(new_terms)
  new_term_ids <- list_drop_empty(new_term_ids)
  # Concatenate new lists on ";" delimiter to create curated value
  sex_cols_cbio$curated_sex[x] <- rmv_xtra_semis(paste(as.list(unique(new_terms)), collapse= ";"))
  sex_cols_cbio$curated_sex_ontology_term_id[x] <- rmv_xtra_semis(paste(as.list(unique(new_term_ids)), collapse= ";"))
  if(x %% 10000==0){print(x)}
}
```


## Curated Table Creation

Next, we can clean up the dataframe by updating column names as needed:
```{r}
# Create a column of relevant source columns
for (i in 1:nrow(sex_cols_cbio)){
  sex_cols_cbio$curated_sex_source[i] <- paste(colnames(sex_cols_cbio[,2:18][which(!is.na(sex_cols_cbio[i,2:18]))]), collapse=";")
}

# Create a curated dataframe
curated_sex <- sex_cols_cbio[,c(1, 19:22)]

# Rename the columns for accuracy & specificity
colnames(curated_sex)[2] <- "original_sex"

# Replace empty values from curated columns with "NA"
curated_sex <- data.frame(lapply(curated_sex, function(x) gsub("^$", NA, x)))
```


## Sanity Check

Before we export this curation, we will conduct a sanity check of sex values
across all patient ids to ensure that they contain the same value. Many patients
included in this curation contributed multiple samples that were utilized across
different studies, and this step will help to validate these curated values.

First we will create a dataframe called `chk` which summarizes all patients with
conflicting values for curated_sex across their samples:
```{r}
# Create a dataframe of curated_sex with a patient_id column
patient_chk <- data.frame(curated_sex, patient_id = cbio$patientId)

# Create a datframe summarizing all patients with more than 1 value listed for curated_sex
chk <- patient_chk %>% 
    group_by(patient_id) %>%
    # Create a column of the number of samples for each patient_id
    summarise(sampleSize = n(),
              uniqueVal = length(unique(curated_sex[!is.na(curated_sex)]))) %>%
    filter(uniqueVal > 1)
```

Next we will expand the `chk` dataframe to generate a column of "best" curated_sex
values based on the sex which is most common across the patient's samples.
Additionally we will track the number of dissenting values and the studies from
which they originated, as well as any patients with no clear "majority" value 
for curated_sex:
```{r}
# Add summary columns to chk dataframe to identify majority and dissenting values for curated_sex
for (pat_id in chk$patient_id){
  # Count the number of "Male" values for the patient_id
  chk$num_male[which(chk$patient_id==pat_id)] <- length(which(patient_chk$patient_id==pat_id & patient_chk$curated_sex=="Male"))
  
  # Count the number of "Female" values for the patient_id
  chk$num_female[which(chk$patient_id==pat_id)] <- length(which(patient_chk$patient_id==pat_id & patient_chk$curated_sex=="Female"))
  
  # Count the number of NA values for the patient_id
  chk$num_na[which(chk$patient_id==pat_id)] <- length(which(patient_chk$patient_id==pat_id & is.na(patient_chk$curated_sex)))
  
  # Set best_value and list studies associated with dissenting value
  if (chk$num_male[which(chk$patient_id==pat_id)] > chk$num_female[which(chk$patient_id==pat_id)]){
    chk$best_value[which(chk$patient_id==pat_id)] <- "Male"
    chk$minority_studies[which(chk$patient_id==pat_id)] <- paste(unlist(lapply(patient_chk$curation_id[which(patient_chk$patient_id==pat_id & patient_chk$curated_sex!="Male")], function(x)sub(":.*", "", x))), collapse=";")
  }else if(chk$num_male[which(chk$patient_id==pat_id)] < chk$num_female[which(chk$patient_id==pat_id)]){
    chk$best_value[which(chk$patient_id==pat_id)] <- "Female"
    chk$minority_studies[which(chk$patient_id==pat_id)] <- paste(unlist(lapply(patient_chk$curation_id[which(patient_chk$patient_id==pat_id & patient_chk$curated_sex!="Female")], function(x)sub(":.*", "", x))), collapse=";")
  }else{
    chk$best_value[which(chk$patient_id==pat_id)] <- NA
    chk$minority_studies[which(chk$patient_id==pat_id)] <- NA
  }
}

# Export "chk" dataframe as a csv file
write.csv(chk, file = file.path(proj_dir, "conflicting_patients.csv"), 
          row.names=F)
```

Next we will make a dataframe summarizing the "problem" studies to determine how
many of the dissenting values originated from each study:
```{r}
# Create a datframe summarizing all unique minority_study values
prob_studies <- chk %>% 
    group_by(minority_studies) %>%
    # Create a column of the number of times each study occurs
    summarise(frequency = n())
```

After reviewing the problem studies, it is apparent that a significant curation
error has affected the `luad_msk_npjpo_2021` study, as 306 of the curated_sex 
values originating from this study are in dissention from the majority value
of those patients. 

We can create a dataframe of the remaining 119 patient IDs from the 
`luad_msk_npjpo_2021` study to investigate if all values from the study are 
in disagreement on curated_sex or if only some values are in disagreement. To do
this, we will iterate through these remaining patient IDs to find the number of
additional studies in which each patient ID is present; if it is not present in 
any additional studies then we cannot confirm if the curated_sex value is accurate,
however if it is present in additional studies then we know the curated_sex value 
is accurate because the patient ID was excluded from the original dataframe of
conflicting values:
```{r}
# Create a dataframe of patient_id values which did not have cross-study curated_sex conflicts
luad_oops <- cbio %>% select(patientId, studyId, SEX) %>% 
  filter(studyId=="luad_msk_npjpo_2021", !(patientId %in% chk$patient_id))

# Confirm that these patients were featured in multiple other studies
luad_oops$other_studies <- sapply(luad_oops$patientId, function(x)length(grep(x, curated_sex$curation_id[which(cbio$studyId!="luad_msk_npjpo_2021")])))
```

Every patient ID was present in at least one additional study, in which the 
curated_sex value was in agreement with the value listed for the patient in 
the `luad_msk_npjpo_2021` study. Therefore we can conclude that the "sex" values
for the entire study were not input incorrectly, and more likely only the 
previously identified values were incorrect.

Additionally, after email correspondence with the curation team at cBioPortal,
we have received confirmation that these mismatched sex values in the 
luad_msk_npjpo_2021 study were curation errors. Therefore we can now replace
the mismatched values for each patient with the "best value" from the chk 
dataframe:

```{r}
# Standardize sex values across all samples for each patient
for (pat in chk$patient_id){
  # Set all sex values for a patient to be the majority value
  curated_sex$curated_sex[which(cbio$patientId==pat)] <- chk$best_value[which(chk$patient_id==pat)]
  # Set all curated ontology term ids to match the patient's curated sex value
  if (is.na(chk$best_value[which(chk$patient_id==pat)])){
    curated_sex$curated_sex_ontology_term_id[which(cbio$patientId==pat)] <- NA
  }else{
    curated_sex$curated_sex_ontology_term_id[which(cbio$patientId==pat)] <- mapped_terms$curated_ontology_term_id[which(mapped_terms$curated_ontology==chk$best_value[which(chk$patient_id==pat)])]
  }
}
```

Last, we will identify any patient ids with one NA value and one descriptive 
value, and fill in the missing value with the descriptive one:
```{r}
# Create a datframe summarizing all patients with more than 1 value listed for curated_sex
chk1 <- patient_chk %>% 
    group_by(patient_id) %>%
    # Create a column of the number of samples for each patient_id
    summarise(sampleSize = n(),
              uniqueVal = length(unique(curated_sex[!is.na(curated_sex)]))) %>%
    filter(uniqueVal == 1) %>% filter(sampleSize > 1)

# For each patient id, find the descriptive value
for (pat in chk1$patient_id){
  # Find the standard value for the patient
  standard_value <- unlist(unique(curated_sex$curated_sex[which((cbio$patientId==pat) & !is.na(curated_sex$curated_sex))]))
  # Find the standard id for the patient
  standard_id <- unlist(unique(curated_sex$curated_sex_ontology_term_id[which((cbio$patientId==pat) & !is.na(curated_sex$curated_sex_ontology_term_id))]))
  # Set all values for the patient to be the descriptive value
  curated_sex$curated_sex[which(cbio$patientId==pat)] <- standard_value
  # Set all curated ontology term ids to match the patient's standardized value
  curated_sex$curated_sex_ontology_term_id[which(cbio$patientId==pat)] <- standard_id
}
```

## Export

Finally, we will export our completed table to GitHub.
```{r export_curated_table, eval=FALSE}
# export to GitHub
write.csv(curated_sex, file = file.path(proj_dir, "curated_sex.csv"), 
          row.names=F)
```


# Update
```{r}
datDir <- "~/OmicsMLRepo/OmicsMLRepoData/cBioPortalData/data"
sex <- readr::read_csv(file.path(datDir, "curated_sex.csv"))
```

```{r}
disease <- readr::read_csv(file.path(datDir, "curated_disease.csv"))
ref <- disease %>% 
    select(curated_cancer_type) %>% 
    filter(!duplicated(curated_cancer_type))
```

## Export for manual harmonization
```{r eval=FALSE}
## Export the template to Google Drive
url <- "https://docs.google.com/spreadsheets/d/1O_W_QDUZKWRNy4GbDQp3RCObb0FCBSVZPrkQA-0zdPk/edit?usp=sharing"
ss <- googledrive::as_id(url)
googlesheets4::write_sheet(ref, ss = ss, sheet = "cBioPortal_sex_inference")
```

Perform manual harmonization and import it back:
```{r}
url <- "https://docs.google.com/spreadsheets/d/1O_W_QDUZKWRNy4GbDQp3RCObb0FCBSVZPrkQA-0zdPk/edit?usp=sharing"
ss <- googledrive::as_id(url)
sex_map <- googlesheets4::read_sheet(ss = ss, sheet = "cBioPortal_sex_inference")

## Subset the ones to update
ind <- which(rowSums(!is.na(sex_map[2:3])) == 2)
map <- sex_map[ind,]
```

Summary of the inaccurate sex record based on the cancer type
```{r}
tb <- left_join(sex, disease[,c("curation_id", "curated_cancer_type")], by = "curation_id")
res <- tb %>% 
    filter(curated_cancer_type %in% map$curated_cancer_type) %>%
    group_by(curated_cancer_type) %>%
    summarise(female = sum(curated_sex == "Female", na.rm = TRUE),
              male = sum(curated_sex == "Male", na.rm = TRUE),
              na = sum(is.na(curated_sex)),
              n = n())
res
```

```{r}
updated_tb <- tb

for (i in seq_len(nrow(map))) {
    
    ## Correct information
    ct <- map$curated_cancer_type[i] # cancer_type
    ct_sex <- map$infered_sex[i] # sex infered from cancer_type
    ct_sex_id <- map$infered_sex_ontology_term_id[i]
    
    ## Identify the wrong entry
    sub <- tb %>%
        filter(curated_cancer_type == ct) %>%
        filter(curated_sex != ct_sex | is.na(curated_sex)) %>%
        pull(curation_id)
    
    ## Fixing
    ind <- which(updated_tb$curation_id %in% sub)
    updated_tb[ind, "curated_sex"] <- ct_sex
    updated_tb[ind, "curated_sex_ontology_term_id"] <- ct_sex_id
    updated_tb[ind, "curated_sex_source"] <- "curated_cancer_type"
}
```

```{r eval=FALSE}
write.csv(updated_tb[,1:5],
          file.path(datDir, "curated_sex.csv"), 
          row.names = FALSE)
```

