---
title: "Untitled"
output: html_document
date: "2024-11-13"
---

# Load the data
We think 'ancestry' is the better concept to describe biological/cultural 
origin of a person; however, most of data schema, including NIH endorsed
CDE, adopt both race and ethnicity. To be compatible with them, we decide
to re-record the harmonized ancestry information into those two attributes.

```{r}
dat <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/cBioPortalData/data/curated_population_ancestry.csv")
```

```{r}
map <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/cBioPortalData/maps/cBioPortal_population_ancestry_detailed_map.csv")
```


# Get dynamic enums
We find the dynamic enums to separate curated terms into 'Race' and 'Ethnicity'.

```{r}
v1 <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/cBioPortal_curated_metadata_release_v1.csv")
v2 <- readr::read_csv("~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/cBioPortal_curated_metadata_release_v2.csv")
```

```{r}
v1All <- v1 %>%
    select(pop_ancestry_ontology_term_id) %>%
    tidyr::separate_longer_delim(., colnames(.), ";") %>%
    tidyr::separate_longer_delim(., colnames(.), "<;>") %>%
    tidyr::separate_longer_delim(., colnames(.), "::") %>%
    unique() %>%
    na.omit() %>%
    pull(pop_ancestry_ontology_term_id)
length(v1All)
```

```{r}
v2AllMain <- v2 %>%
    select(population_ancestry_ontology_term_id) %>%
    tidyr::separate_longer_delim(., colnames(.), ";") %>%
    tidyr::separate_longer_delim(., colnames(.), "<;>") %>%
    tidyr::separate_longer_delim(., colnames(.), "::") %>%
    unique() %>%
    na.omit() %>%
    pull(population_ancestry_ontology_term_id)
v2AllDetail <- v2 %>%
    select(population_ancestry_detailed_ontology_term_id) %>%
    tidyr::separate_longer_delim(., colnames(.), ";") %>%
    tidyr::separate_longer_delim(., colnames(.), "<;>") %>%
    tidyr::separate_longer_delim(., colnames(.), "::") %>%
    unique() %>%
    na.omit() %>%
    pull(population_ancestry_detailed_ontology_term_id)

v2All <- unique(c(v2AllMain, v2AllDetail))
length(v2All)

setdiff(v1All, v2All)
```

```{r}
dynamic_enums <- OmicsMLRepoCuration::mapNodes(allTerms, cutoff = 0.9)
```

