---
title: "Harmonize study design and target phenotype information in cbio data"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```


# Overview

This .qmd file demonstrates a workflow for curating and harmonizing 
study design and target phenotype data with the cbioportal dataset.
This file curates information derived from the study descriptions to produce a 
curated column describing the study design format as well as a curated column
describing the disease and/or condition(s) focused on in each study.


# Setup

## Load Packages

```{r load}
suppressPackageStartupMessages({
 library(dplyr)
library(tidyverse)
library(stringr)
library(rvest)
library(rols)
library(vctrs)
library(hash)
})
```


## Setup for Curation

We will first add a `curation_id` consisting of study ID and patient ID (`studyId:patientId`) to avoid confusion due to duplicated samples.
```{r curation_setup}
# Load cbio data
cbio <- readRDS("C:\\Users\\Owner\\Desktop\\CUNY Internship\\cBioPortal_all_clinicalData_combined_2022-10-14.rds")

# Add curation ID
cbio$curation_id <- paste(cbio$studyId, cbio$patientId, sep = ":")

# Setup google sheets link to mapping table
url <- "https://docs.google.com/spreadsheets/d/1IgrVEdgCZdvBmWrER21A57lSkfDjdRdV3RbK_yoqMl4/edit?usp=sharing"

ss <- googledrive::as_id(url)

# Define a project directory
proj_dir <- "C:\\Users\\Owner\\Desktop\\CUNY Internship\\cbio\\StudyDesign\\"
```


# Exploratory Data Analysis & Data Curation

## Identifying Relevant Columns

Next we will create a new dataframe containing only the columns that are relevant
to this curation, which are the curation_id and studyId columns. 
```{r}
# Create tibble
cbio <- tibble(cbio)

# Create a subset of cbio containing only relevant columns
studydesign_cols_cbio <- cbio[,c("curation_id", "studyId")]
```

Then we will create a dataframe of only the unique studyId values, and export 
this dataframe to a csv file for manual development of an ontology map of the 
corresponding study designs and target phenotypes:
```{r}
# Create a summary dataframe of the relevant columns
unique_studies <- data.frame("study_id"=unique(studydesign_cols_cbio$studyId))

# Create a csv file for manual curation
write.csv(unique_studies, 
          file = file.path(proj_dir, "unique_study_ids.csv"),
          row.names = FALSE)
```


# Mapping

## Import Maps

To access the necessary study design and target phenotype ontology terms we will 
load the `study_design_ontology_map` ontology map file.
```{r import_maps, eval=FALSE}
# import bodysite ontology map
mapped_terms <- read_sheet(ss, sheet = "study_design_ontology_map")

# import bodysite ontology map
mapped_terms <-read.csv(file = file.path(proj_dir, "cBioPortal_study_design_ontology_map.csv"), header=T)
```


# Curating Study Design Columns

Now we can map the study_id values in the `studydesign_cols_cbio` dataframe to 
their corresponding study_design and target_phenotype terms and propagate these 
values into two new curated columns: 
```{r}
# Format columns for mapping
colnames(studydesign_cols_cbio)[2] <- "study_id"
studydesign_cols_cbio$target_phenotype_term_id <-
studydesign_cols_cbio$target_phenotype <-
studydesign_cols_cbio$study_design <- NA

# Iterate through merged column values
for(x in 1:length(studydesign_cols_cbio$study_id)){
  studydesign_cols_cbio$study_design[x] <- paste(mapped_terms$study_design[which(studydesign_cols_cbio$study_id[x]==mapped_terms$study_id)])
  studydesign_cols_cbio$target_phenotype[x] <- paste(mapped_terms$target_phenotype[which(studydesign_cols_cbio$study_id[x]==mapped_terms$study_id)])
  studydesign_cols_cbio$target_phenotype_term_id[x] <- paste(mapped_terms$target_phenotype_term_id[which(studydesign_cols_cbio$study_id[x]==mapped_terms$study_id)])
  # Processing Checkpoints
  if(x %% 10000==0){print(x)}
}
```


## Export

Finally, we will export our completed table to GitHub.
```{r export_curated_table, eval=FALSE}
# export to GitHub
write.csv(studydesign_cols_cbio, file = file.path(proj_dir, "curated_study_design.csv"), 
          row.names=F)
```