---
title: "Harmonize feces phenotype information in cMD sample metadata"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```

# Overview

This .qmd file demonstrates a workflow for curating and harmonizing patient
data derived from fecal samples with the curatedMetagenomicData package.
This file curates information from the original bristol_score, HBI, and 
calprotectin columns to produce the curated feces_phenotype column.

# Setup
## Load Packages

```{r load}
suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(tidyverse)
  library(googlesheets4)
  library(rols)
  library(dplyr)
  library(hash)
  library(vctrs)
  library(rvest)
})
```

## Setup for Curation

We will first add a `curation_id` consisting of study name and sample ID (`study_name:sample_id`) to avoid confusion due to duplicated samples.

```{r curation_setup}
# Add curation ID
sampleMetadata$curation_id <- paste(sampleMetadata$study_name, sampleMetadata$sample_id, sep = ":")

# Define a project directory
proj_dir <- "C:\\Users\\Owner\\Desktop\\CUNY Internship\\cMD"
```

# Exploratory Data Analysis & Data Curation

## Manual Data Corrections

Next, we will make some manual corrections to the data.
We will start by creating a tibble dataframe and checking the completeness of 
the three original columns as well as reviewing their unique values:
```{r}
# Create a tibble object 
samplemd <- as_tibble(sampleMetadata)

# Review unique values
unique(sampleMetadata$bristol_score)
unique(sampleMetadata$calprotectin)
unique(sampleMetadata$HBI)

# Calculate completeness
length(samplemd$bristol_score[which(!is.na(samplemd$bristol_score))])/length(samplemd$bristol_score)
length(samplemd$calprotectin[which(!is.na(samplemd$calprotectin))])/length(samplemd$calprotectin)
length(samplemd$HBI[which(!is.na(samplemd$HBI))])/length(samplemd$HBI)
```

Then we can create a list of the names of the columns which will be included in 
the merged column,and export these values to a csv file to manually develop an 
ontology map:
```{r}
# Create a dataframe of only the relevant columns
fecal_cols <- samplemd %>% select(curation_id, bristol_score, calprotectin, HBI)
fecal_cols <- as.data.frame(sapply( fecal_cols, as.character))

# Get column names for ontology map
unique_merged_cols <- colnames(fecal_cols[2:4])

# Make a csv file of the unique values in the merged column
write.csv(unique_merged_cols, 
          file= file.path(proj_dir, "fecal_cols_unique_vals.csv"), 
          row.names=F, col.names=F)
```


# Mapping

## Import Maps

To access the necessary feces phenotype ontology terms we will load the 
`cMD_feces_phenotype_map`.
```{r import_maps, eval=FALSE}
# import smoker ontology map
# mapped_terms <- read_sheet(ss, sheet = "feces_phenotype_ontology")

# import smoker ontology map
mapped_terms <-read.csv(file = file.path(proj_dir, "cMD_feces_phenotype_map.csv"), header=T)
```

# Curating Feces_Phenotype Column

Next we can convert plain numeric values into a combined descriptive and numeric 
value pair by adding the corresponding column name. We will also take this 
opportunity to map the new values associated with each column metric and their
corresponding ontology term IDs:
```{r}
# Initiate an ontology_term_id column
fecal_cols$curated_feces_phenotype_ontology_term_id <- "8X8"

# Merge all relevant rows into an original_feces_phenotype column
for (i in 1:nrow(fecal_cols)){
  fecal_cols$original_feces_phenotype[i] <- toupper(paste(fecal_cols[i,2:4], collapse=";"))
}
# Take out NA values
fecal_cols$original_feces_phenotype <- gsub("NA;", "", fecal_cols$original_feces_phenotype)
fecal_cols$original_feces_phenotype <- gsub(";NA", "", fecal_cols$original_feces_phenotype)
fecal_cols$original_feces_phenotype <- gsub("NA", NA, fecal_cols$original_feces_phenotype)


# Update numeric columns with additional ontology values for metric type
for(col in colnames(fecal_cols[,2:4])){
  new_col <- mapped_terms$curated_ontology[grep(paste("^",col,"$",sep=""), mapped_terms$original_value, fixed=F)]
  new_col_onto <- mapped_terms$curated_ontology_term_id[grep(paste("^",col,"$",sep=""), mapped_terms$original_value, fixed=F)]
  for (row in 1:nrow(fecal_cols)){
    val <- fecal_cols[row,col]
    if (!is.na(val)){
      fecal_cols[row,col] <- paste(new_col, ":", fecal_cols[row,col], sep="")
      fecal_cols$curated_feces_phenotype_ontology_term_id[row] <- 
        paste(fecal_cols$curated_feces_phenotype_ontology_term_id[row], new_col_onto, sep=";")
    }
  }
}
```

Now we can replace character values representing a lack of information with a 
unique character value ("8X8"), and merge these values into a single column with 
each distinct value separated by a semicolon.
```{r}
# Convert NA values to unique character value (8X8)
fecal_cols[,2:4] <- fecal_cols[,2:4] %>% replace(is.na(.), "8X8") %>% data.frame()

# Merge all relevant rows
for (i in 1:nrow(fecal_cols)){
  fecal_cols$curated_feces_phenotype[i] <- paste(fecal_cols[i,2:4], collapse=";")
}
```

Next, we can remove the unique character representing NA values from the merged
column:
```{r}
# Take out NA values
fecal_cols$curated_feces_phenotype <- gsub("8X8;", "", fecal_cols$curated_feces_phenotype)
fecal_cols$curated_feces_phenotype <- gsub(";8X8", "", fecal_cols$curated_feces_phenotype)
fecal_cols$curated_feces_phenotype_ontology_term_id <- 
  gsub("8X8;", "", fecal_cols$curated_feces_phenotype_ontology_term_id)
```


## Curated Table Creation

Next, we can clean up the dataframe by replacing the unique character 
representing NA values with "NA" and updating column names as needed:
```{r}
# Replace NA values from original columns with "NA"
fecal_cols <- data.frame(lapply(fecal_cols, 
                                          function(x) gsub("8X8", NA, x)))

# Create a column of relevant source columns
for (i in 1:nrow(fecal_cols)){
  fecal_cols$curated_feces_phenotype_source[i] <- paste(colnames(fecal_cols[,2:4][which(!is.na(fecal_cols[i,2:4]))]), collapse=";")
}

# Create a curated dataframe
curated_feces_phenotype <- fecal_cols[,c(1, 6, 7, 5, 8)]

# Replace empty values from curated columns with "NA"
curated_feces_phenotype <- data.frame(lapply(curated_feces_phenotype, 
                                      function(x) gsub("^$", NA, x)))
```


## Export

Finally, we will export our completed table to GitHub.
```{r export_curated_table, eval=FALSE}
# export to GitHub
write.csv(curated_feces_phenotype, 
          file = file.path(proj_dir, "curated_feces_phenotype.csv"),
          row.names = FALSE)
```