---
title: "Harmonize FMT information in cMD sample metadata"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```

# Overview

This .qmd file demonstrates a workflow for curating and harmonizing FMT data 
with the curatedMetagenomicData package.
This file curates information from the original fmt_id and FMT_role columns to 
produce the curated FMT_id and FMT_role columns.

# Setup
## Load Packages

```{r load}
suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(tidyverse)
  library(googlesheets4)
  library(rols)
  library(dplyr)
  library(hash)
  library(vctrs)
  library(rvest)
})
```

## Setup for Curation

We will first add a `curation_id` consisting of study name and sample ID (`study_name:sample_id`) to avoid confusion due to duplicated samples.

```{r curation_setup}
# Add curation ID
sampleMetadata$curation_id <- paste(sampleMetadata$study_name, sampleMetadata$sample_id, sep = ":")

# Define a project directory
proj_dir <- "C:\\Users\\Owner\\Desktop\\CUNY Internship\\cMD"
```

# Exploratory Data Analysis & Data Curation

## Manual Data Corrections

Next, we will make some manual corrections to the data.
We will start by creating a tibble dataframe and checking the completeness of 
the two original columns as well as reviewing their unique values:
```{r}
# Create a tibble object 
samplemd <- as_tibble(sampleMetadata)

# Review unique values
unique(sampleMetadata$fmt_id)
unique(sampleMetadata$FMT_role)

# Calculate completeness
length(samplemd$fmt_id[which(!is.na(samplemd$fmt_id))])/length(samplemd$fmt_id)
length(samplemd$FMT_role[which(!is.na(samplemd$FMT_role))])/length(samplemd$FMT_role)
```

Now we can replace the miss-typed value "recipient-before" with the correct
format "recipient_before" and rename the `fmt_id` column to `FMT_id`.
```{r}
# Subset fmt cols
fmt_cols <- samplemd %>% select(curation_id, fmt_id, FMT_role)

# Fix the typo in the FMT_role column
fmt_cols$curated_FMT_role <- gsub("recipient-before", "recipient_before", fmt_cols$FMT_role)
unique(fmt_cols$curated_FMT_role)

# Change column name 
fmt_cols <- fmt_cols %>% rename(curated_FMT_id = "fmt_id", original_FMT_role = "FMT_role")

```


## Export

Finally, we will export our completed table to GitHub.
```{r export_curated_table, eval=FALSE}
# export to GitHub
write.csv(fmt_cols, 
          file = file.path(proj_dir, "curated_FMT.csv"),
          row.names = FALSE)
```