---
title: "Harmonize population ancestry information in cMD sample metadata"
author:
  - Sehyun Oh
  - Kai Gravel-Pucillo
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                    warning = FALSE,
                    message = FALSE,
                    collapse = TRUE,
                    eval = TRUE)
```

# Overview

This .qmd file demonstrates a workflow for curating and harmonizing population 
ancestry data with the curatedMetagenomicData package.
This file curates information from the original population column to 
produce the curated population ancestry column.

# Setup
## Load Packages

```{r load}
suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(tidyverse)
  library(googlesheets4)
  library(rols)
  library(dplyr)
  library(hash)
  library(vctrs)
  library(rvest)
})
```

## Setup for Curation

We will first add a `curation_id` consisting of study name and sample ID (`study_name:sample_id`) to avoid confusion due to duplicated samples.

```{r curation_setup}
# Add curation ID
sampleMetadata$curation_id <- paste(sampleMetadata$study_name, sampleMetadata$sample_id, sep = ":")

# Define a project directory
proj_dir <- "D:\\CUNY Internship\\cMD"
```

# Exploratory Data Analysis & Data Curation

## Manual Data Corrections

Next, we will make some manual corrections to the data.
We will start by creating a tibble dataframe and checking the completeness of 
the original column as well as reviewing the unique values:
```{r}
# Create a tibble object 
samplemd <- as_tibble(sampleMetadata)

# Review unique values
unique(sampleMetadata$population)

# Calculate completeness
length(samplemd$population[which(!is.na(samplemd$population))])/length(samplemd$population)
```

Now we can replace character values representing a lack of information with a 
unique character value ("8X8"), and merge these values into a single column with 
each distinct value separated by a semicolon.
```{r}
# Create a dataframe of only the relevant columns
population_cols <- samplemd %>% select(curation_id, population)

# Convert NA values to unique character value (8X8)
population_cols <- population_cols %>% replace(is.na(.), "8X8") %>% data.frame()

# Merge all relevant rows
for (i in 1:nrow(population_cols)){
  population_cols$merged[i] <- toupper(paste(population_cols[i,2], collapse=";"))
}
```

Next, we can remove the unique character representing NA values from the merged
column:
```{r}
# Take out NA values
population_cols$merged <- gsub("8X8;", "", population_cols$merged)
population_cols$merged <- gsub(";8X8", "", population_cols$merged)
```

Then we can create a list of all unique values included in the merged column,
and export these values to a csv file to manually develop an ontology map:
```{r}
# Count the number of unique values in the merged column
merged_vals <- list()
for(r in 1:nrow(population_cols)){
  merged_vals <- append(merged_vals, str_split(population_cols$merged[r], ";"))
}
unique_merged_vals <- unique(unlist(merged_vals))

# Make a csv file of the unique values in the merged column
write.csv(unique_merged_vals, 
          file= file.path(proj_dir, "population_cols_unique_vals.csv"), 
          row.names=F, col.names=F)
```

# Mapping

## Import Maps

To access the necessary population ontology terms we will load the 
`population_ontology_map`.
```{r import_maps, eval=FALSE}
# import population ontology map
# mapped_terms <- read_sheet(ss, sheet = "population_map")

# import population ontology map
mapped_terms <-read.csv(file = file.path(proj_dir, "cMD_population_ancestry_map.csv"), header=T)
```


# Curating Population Ancestry Column

Now we can map the values in the merged columns to their corresponding ontology
terms and propagate these values into the curated columns: 
```{r}
# Make a function to remove excess semicolons
rmv_xtra_semis <- function(x){
  x <- ifelse(startsWith(x, ";"), sub("^;", "", x), x)
  return(ifelse(endsWith(x, ";"), sub(";$", "", x), x))
}

# Iterate through merged column values
for (x in 1:length(population_cols$merged)){
  # Create a list of terms in the value
  original_terms <- as.list(unlist(strsplit(population_cols$merged[x], ";")))
  new_terms <- list()
  new_term_ids <- list()
  # Search for replacement terms in the ontology map
  new_terms <- lapply(original_terms, function(y) mapped_terms$curated_ontology[grep(paste("^",y,"$",sep=""), mapped_terms$original_value, fixed=F)])
  new_term_ids <- lapply(original_terms, function(y) mapped_terms$curated_ontology_term_id[grep(paste("^",y,"$",sep=""), mapped_terms$original_value, fixed=F)])
  new_terms <- list_drop_empty(new_terms)
  new_term_ids <- list_drop_empty(new_term_ids)
  # Concatenate new lists on ";" delimiter to create curated value
  population_cols$curated_population_ancestry[x] <- rmv_xtra_semis(paste(as.list(unique(new_terms)), collapse= ";"))
  population_cols$curated_population_ancestry_ontology_term_id[x] <- rmv_xtra_semis(paste(as.list(unique(new_term_ids)), collapse= ";"))
  if(x %% 10000==0){print(x)}
}
```


## Curated Table Creation

Next, we can clean up the dataframe by replacing the unique character 
representing NA values with "NA" and updating column names as needed:
```{r}
# Replace NA values from original columns with "NA"
population_cols <- data.frame(lapply(population_cols, 
                                          function(x) gsub("8X8", NA, x)))

# Create a column of relevant source columns
for (i in 1:nrow(population_cols)){
  population_cols$curated_population_ancestry_source[i] <- ifelse(is.na(population_cols[i,2]), NA, colnames(population_cols)[2])
}

# Create a curated dataframe
curated_population <- population_cols[,c(1, 3:6)]

# Rename the columns for accuracy & specificity
colnames(curated_population)[2] <- "original_population_ancestry"

# Replace empty values from curated columns with "NA"
curated_population <- data.frame(lapply(curated_population, 
                                      function(x) gsub("^$", NA, x)))
```


## Export

Finally, we will export our completed table to GitHub.
```{r export_curated_table, eval=FALSE}
# export to GitHub
write.csv(curated_population, 
          file = file.path(proj_dir, "curated_population_ancestry.csv"),
          row.names = FALSE)
```