---
title: "Harmonize location/period information in cMD sample metadata"
author: "Sehyun Oh"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14px
        toc: true
        top-depth: 3
abstract: "Prepare U24 Supplement: AI/ML-ready"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE, 
                      collapse = TRUE, 
                      eval = FALSE)
```

# Load metadata
```{r}
dir <- "~/OmicsMLRepo/OmicsMLRepoData/inst/extdata/"
cmd <- read.csv(file.path(dir, "cMD_curated_sampleMetadata.csv"))
cbio <- readRDS(file.path(dir, "cBioPortal_all_clinicalData_combined_2022-10-14.rds"))
```

# Geographical Location
## cMD
`country` and `location` columns seems relevant. (`country` using ISO3 codes.)
```{r}
ind <- grep("location|country|city", colnames(cmd))
apply(cmd[,ind], 2, function(x) {head(unique(x))})
```

```{r}
head(sort(table(cmd$location), decreasing = TRUE))
OmicsMLRepoCuration::calCompleteness(c("country", "location"), DB = "cMD")
```


## cBioPortal
`COUNTRY`, `COUNTRY_OF_ORIGIN` (more for ethnicity?), `CITY_OF_PROCUREMENT`, 
`COUNTRY_OF_PROCUREMENT`, `TUMOR_SAMPLE_PROCUREMENT_COUNTRY`, `BIRTH_COUNTRY`,
`PARTICIPANT_COUNTRY`, `TX_LOCATION`.
```{r eval=FALSE}
ind <- grep("location|country|city", tolower(colnames(cbio)))
apply(cbio[,ind], 2, function(x) {head(unique(x))})
```

```{r}
cbio_locCol <- c("COUNTRY", "COUNTRY_OF_ORIGIN", "CITY_OF_PROCUREMENT",
                 "COUNTRY_OF_PROCUREMENT", "TUMOR_SAMPLE_PROCUREMENT_COUNTRY",
                 "BIRTH_COUNTRY", "PARTICIPANT_COUNTRY", "TX_LOCATION")

OmicsMLRepoCuration::calCompleteness(cbio_locCol, DB = cbio)
num_loc_available <- rowSums(!is.na(cbio[,cbio_locCol]))
table(num_loc_available)
```

```{r}
nonNA_ind <- which(num_loc_available != 0) # any location info
head(cbio[nonNA_ind, cbio_locCol])

nonNA_ind <- which(rowSums(!is.na(cbio[,cbio_locCol])) > 1) # >=2 location info
head(cbio[nonNA_ind, cbio_locCol])
```


# Recruitment Period
Very little information on recruitment period overall, so I think it might not
be worth to harmonize this. --> Keep this this attribute on hold.

## cMD
```{r eval=FALSE}
ind <- grep("date|time|period|year", colnames(cmd))
apply(cmd[,ind], 2, function(x) {head(unique(x))})
```

## cBioPortal
```{r eval=FALSE}
ind <- grep("date|time|period|year", tolower(colnames(cbio)))
apply(cbio[,ind], 2, function(x) {head(unique(x))})
```

```{r eval=FALSE}
cbio_timeCol <- c("FORM_COMPLETION_DATE", "INITIAL_PATHOLOGIC_DX_YEAR",
                  "AGE_AT_SEQ_REPORTED_YEARS", "YEAR_OF_DIAGNOSIS",
                  "DATE_ADDED", "FOLLOW_UP_TIME_MONTHS", 
                  "TIME_TO_LAST_FOLLOWUP", "CRDB_CONSENT_DATE_DAYS",
                  "YEAR_OF_INITIAL_PATHOLOGIC_DIAGNOSIS",
                  "DEATH_DATE", "FOLLOW_UP_DATE", "LAST_FOLLOWUP_YEAR",
                  "YEAR", "YEAR_NAT_START", "YEAR_NAT_START_GROUPS",
                  "YEAR_OF_BIRTH")

OmicsMLRepoCuration::calCompleteness(cbio_timeCol, DB = cbio)

## The number of data points with non-NA value among `cbio_timeCol` columns
num_time_available <- rowSums(!is.na(cbio[,cbio_timeCol]))

nonNA_ind <- which(num_time_available == 5)
head(cbio[nonNA_ind, cbio_timeCol])

apply(cbio[,cbio_timeCol], 2, function(x) {head(unique(x), 10)})
```
