---
title: "cBioPortal"
author: "Sehyun Oh"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cBioPortalData pre/post harmonization stats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Setup
```{r}
suppressPackageStartupMessages({
    library(OmicsMLRepoCuration)
    library(dplyr)
})

devtools::load_all("~/OmicsMLRepo/OmicsMLRepoR")
```

## Load metadata
```{r}
cbio <- getMetadata("cBioPortal")
# colnames(cbio)
```

## Helper function
```{r}
nonNACols <- function(x) {
    colInds <- which(colSums(is.na(x)) < nrow(x))
    res <- x[colInds]
    return(res)
}
```


# Attributes
## Study Sizes
```{r}
x <- cbio %>% group_by(studyId) %>% summarize(study_size = n())
hist(x$study_size, main = "Study Size", xlab = "# samples", ylab = "# studies")
```

## ancestry
```{r eval=FALSE}
ancestry_terms <- cbio %>%
    select(pop_ancestry_ontology_term_id) %>%
    tidyr::separate_longer_delim(., colnames(.), ";") %>%
    tidyr::separate_longer_delim(., colnames(.), "<;>") %>%
    tidyr::separate_longer_delim(., colnames(.), "::") %>%
    unique() %>%
    na.omit() %>%
    .[["pop_ancestry_ontology_term_id"]]

dynamic_enums <- getEnumNodes(ancestry_terms, cutoff = 0.8)
```

## body_site
[Error] The numbers of terms and term_ids are not matching:
```{r error_check_with_Kai, eval=FALSE}
> cbio_body_site_issue_ind <- c(7310, 125029, 125059)
> cbio_body_site[cbio_body_site_issue_ind, c("curated_body_site", "curated_body_site_ontology_term_id")]
# A tibble: 3 × 2
  curated_body_site                                      curated_body_site_ontology_term_id
  <chr>                                                  <chr>                             
1 Cerebellum<;>Posterior Cranial Fossa<;>Brain Ventricle NCIT:C12445<;>NCIT:C33361         
2 Left<;>Temporal Lobe<;>Parietal Lobe                   NCIT:C25229<;>NCIT:C12353         
3 Left<;>Temporal Lobe<;>Parietal Lobe                   NCIT:C25229<;>NCIT:C12353
```

## treatment
Started from the 6 original groups (original group: resultant features)

name: name
type: type
amount: dose_val, dose_unit, number_val, number_unit
time: start_val, start_unit, end_val, end_unit, duration_val, duration_unit, frequency_val, frequency_unit
case: status, reason, group
notes: notes

### treatment name
```{r}
tn <- getLongMetaTb(cbio, targetCols = "treatment_name") 
```

Delimiter usage in `treatment_name`
```{r}
#' Check whether `;` used 
#' 
#' @param meta The metadata table
#' @param attr The name of the attribute to check the existence of the `pattern`
#' @param pattern The regular expression of the pattern to detect. 
#' 
#' @examples
#' pattern <- "(?<!<);(?!>)"  # match ; but not <;>
#' 
checkDelim <- function(meta, attr, pattern = "(?<!<);(?!>)") {
    res <- grep(pattern, meta[[attr]], perl = TRUE)
    length(res)
}
```

```{r}
## The number of rows that contain ; but not <;>
delim_res <- apply(cbio, 2, function(x) grep(pattern, x, perl = TRUE) %>% length())
delim_res[delim_res != 0]
```


```{r}
pattern <- "(?<!<);(?!>)"  # match ; but not <;>
res <- grepl(pattern, cbio$treatment_name, perl = TRUE)
head(cbio$treatment_name[res])
head(cbio[res,])

test <- head(cbio[res,])
# targetCols <- colnames(test)[grep("treatment_", colnames(test))]
targetCols <- paste0("treatment_",
                     c("dose", "duration", "end", "frequency", "group",
                       "name", "notes", "number", "reason", "start", 
                       "status", "type"))
df <- getLongMetaTb(test, targetCols)
```




## Bodysite
```{r}
test <- cbio %>% filter(bodysite == "Soft Tissue<;>Bone<;>Scapula")
nonNACols(test)
```

